Ensemble Stability of Biological Networks

### 2014-09-22 paper.tex
move bounds on robustness section to end
reorganize and add todos

% \section{Todo}
% \listoftodos

@@

\section{Stability analysis of biological networks}\label{sec:stabanalbn}
In the construction of a class of potential models for biological systems at any level of the biological hierarchy from metabolic to ecosystem-level networks, it is common to first attempt to define a collection of observable phenomena of interest and determine a domain (such as binary numbers, integers, or real numbers) in which each observable can be quantified. Next, it is necessary to establish the interdependencies among system components. The specific manner in which the components depend upon one another must be clarified, which is often done by determining a particular system of mathematical functions that represents a hypothesis about how the quantified observables evolve in time. To the degree to which there is uncertainty about the interactions among system components, parameters are introduced to broaden the class of models under consideration. Finally, whatever model class remains may be compared to empirical observations to determine how capable the model is of representing the phenomena of interest.

For the case of continuous deterministic observables, the above process can be made more precise by associating a manifold $M_i$ to each observable, a directed graph, $G$, to the interdependencies, and a vector field $V \in \Gamma T(\prod_i M_i)$ over the space determined by taking the product of the manifolds associated to the collection of observables, satisfying these interdependencies for each observable  \cite{Deville}. For example, if we have two observables $\{x_1,x_2\}$ where the domains in which they are quantified are given by manifolds $\{M_1,M_2\}$ such that $x_1 \in M_1 \equiv \mathbb{R}^1$ and $x_2 \in M_2 \equiv \mathbb{R}^1$, a directed graph $G_X$ describing the dependencies between these observables and vector field $V$ with components $\{F_1,F_2\}$ defined on $M_1 \otimes M_2 \equiv \mathbb{R}^2$ satisfying the dependencies determined by $G_X$. If the system under consideration has the graph given in \ref{fig:examplesystemmodules}A
% \begin{center}
% \input{tex/examplehamdyngraph}
% \end{center}
with adjacency matrix
$$
\adj(G_X) = \begin{pmatrix}
0 & 1 \\
1 & 1
\end{pmatrix}
$$
having connectivity $\connectivity$ equal to the number of edges of the graph, in this case $\connectivity = 3$. For the system $X = \{G_X, M_X, F_X\}$, where $M_X \equiv \{M_1,M_2\}$ and $F_X \equiv \{F_1,F_2\}$ such that $F_1$ is constrained to be a function of $x_2$ and $F_2$ is a function of both $x_1$ and $x_2$ yielding the flow equations
\begin{align*}
\frac{dx_1}{dt} & = F_1(x_2)\\
\frac{dx_2}{dt} & = F_2(x_1,x_2)
\end{align*}

@@

This description can be formalized as a directed graph $G_X$ that describes the manner in which each of the variables depends upon one another given by an adjacency matrix $\adj(G_X)$ where
 \begin{displaymath}
   \adj(G_X)_{ij} = \left\{
     \begin{array}{ll}
       1, & F_i \hbox{ depends on } x_j\\
       0, & F_i \hbox{ does not depend on } x_j
     \end{array}
   \right.
\end{displaymath}
A graph can be decomposed into a natural collection of \emph{modules} called strongly connected components. A strongly connected component of a graph is a maximal subset of vertices where each vertex within the subset can be reached from any other \cite{Cormen2009}. The strongly connected components of some examples of three component systems are outlined in \ref{fig:scc} along with their adjacency matrices and hierarchy diagrams (to be defined below).

Corresponding to this decomposition we can construct a graph $\hier (G)$.  Each node of $\hier (G)$ corresponds to a strongly connected component of $G$. There is an edge from the node corresponding to component $C$ to the node corresponding to component $C'$ iff there exists a link from some vertex in $C$ to some vertex in $C'$ in $G$.  Because of the maximality of strongly connected components, $\hier (G)$ is acyclic.

One can also perform this construction in the oppsite direction.  Start with a directed acyclic graph $A$.  To each node $n$ of $A$ associate
a strongly connected graph $C_n$.  To each link $(i,j)$ of $A$ associate a non-empty subset of $\Vertex(C_i) \times \Vertex(C_j)$.  The result will be a graph $G$ such that $\hier(G) = A$ and furthermore, every graph $G$ such that $\hier(G) = A$ can be obtained in this manner.

Linearization of the $F_i$ about any equilibrium solution $\vec{x}^0$ such that $F_i(\vec{x}^0)=0$ leads to the system
\begin{equation}\label{eq:lineardynsys}
\frac{d\vec{y}(t)}{dt} = A \vec{y}(t),
\end{equation}
where $\vec{y} = \vec{x} - \vec{x}^0$ and the $n \times n$-matrix $A$ has components
$$
a_{ij} = \left. \frac{\partial F_i}{\partial x_j} \right|_{\vec{x} = \vec{x}^0}.
$$
To each dynamical system having Jacobian matrix $A$ at some fixed point $\vec{x}^0$ we can associate a directed graph $G_A$ given by an adjacency matrix $\adj(G_A)$ where
 \begin{displaymath}
   \adj(G_A)_{ij} = \left\{
     \begin{array}{lr}
       1, & a_{ij} \neq 0\\
       0, & a_{ij} = 0
     \end{array}
   \right.
\end{displaymath}
The graphs $G_X$ and $G_A$ are equivalent because the condition $F_i$ independent of $x_j$ definitive of $\adj(G_X)$ corresponds precisely to $\frac{\partial F_i}{\partial x_j}=0$.

@@

\section{Stability and robustness analysis of multi-component systems}
Suppose that $M$ is an $n \times n$ matrix with real coefficients and connectivity $\connectivity$. We will consider the case in which each of the parameters defining the matrix is sampled from the uniform distribution $\mathcal{U}(-1,1)$ so that the parameter space corresponds to the $\connectivity$-dimensional hypercube, $H^\connectivity$, of edge length $r=2$, centered about the origin, and having volume $r^\connectivity$.  We want to know the probability that, if we resample the elements $k_1, k_2, \ldots k_m$ of this matrix, the resulting matrix will correspond to a stable matrix.  This is a special case of \ref{eq:robustness}:
\begin{equation}\label{eq:condprobgen}
 R (\mathrm{stab}, k_1, \ldots k_m) =
  \frac{\int_{H^\connectivity} dx_1 \cdots dx_d \int_{H^m} {dx'}_{k_1} \cdots {dx'}_{k_m}\,
    \mathbf{1}_{stab} \mathbf{1}_{stab'}}
  {\int_{H^\connectivity} dx_1 \cdots dx_d  \, \mathbf{1}_{stab}(x_1 \cdots x_d)}
\end{equation}
where we made the abbreviations
\begin{align*}
\mathbf{1}_{stab} &= \mathbf{1}_{stab}(x_1 \cdots x_{k_1} \cdots x_{k_m} \cdots x_\connectivity) \\
\mathbf{1}_{stab'} &= \mathbf{1}_{stab'}(x_1 \cdots {x'}_{k_1} \cdots {x'}_{k_m}  \cdots x_d)
\end{align*}
i.e. $\mathbf{1}_{stab'}$ corresponds to replacing $x_{k_1}, \ldots x_{k_m}$ with their primed counterparts.

@@

Define the characteristic polynomial as $\chi(M)(z) = \det(zI - M)$.
Denote the coefficients of $\chi(M)$ as $s_j$ and its roots as $r_j$,
writing $p(x) = x^n + \sum_{k=1}^n s_k x^{n-k} = \prod_{k=1}^n (x-r_k)$.
Then, since the roots of $\chi$ are the eigenvalues of $M$, asking that
$M$ be stable is equivalent to asking that the roots of $\chi$ all have
negative real part.  Hence, let us call a polynomial stable if all its
roots have negative real part.  We will formulate conditions for this to
happen in terms of the coefficients of $\chi$ \cite{Gantmacher1959}.

@@

\section{Dependency of robustness upon connectivity and cycle number}
It has been stated previously on the basis of simulation that system stability decreases with connectivity as the system size goes to infinity \cite{May1972}. For small system sizes such as the two and three component systems, studied here the situation is not so clear cut. For two component systems, system stability is constant across the entire range of connectivities. For three component systems, the trend shows a minor decrease from connectivity $4$ to $5$ followed by small fluctuations as shown in \ref{fig:combined}A.
The relationship between connectivity and robustness for two component systems is shown in \ref{tab:structstabmat} and likewise for three component systems in \ref{tab:structstabmat3} and \ref{fig:combined}B. If we average over the different classes of matrices for a given connectivity we see there is a correlation between connectivity and robustness demonstrated by the red dots in \ref{fig:combined}B.
% \section{Cycle number is inversely correlated with robustness}
\ref{fig:combined}C shows the robustness for all three component systems as a function of the number of simple cycles (elementary circuits) of length greater than one in the corresponding directed graph \cite{Johnson1975}. There appears to be a weak negative correlation between robustness and the number of simple cycles.

@@

\section{Component decomposition and robustness}

@@

Next, we ask what is the maximally robust graph that has a given set of connected components.
\begin{align}
R_{\mathrm{max}} = \frac{\frac{1}{2} ((\sum_i n_i)^2 - \sum_i n_i^2) + \sum_i c_i R_i}{\frac{1}{2} ((\sum_i n_i)^2 - \sum_i n_i^2) + \sum_i c_i}, \label{eq:sccmaxrobustness}
\end{align}

\section{Strongly connected components, robustness, and hierarchy}
The fact that the most hierarchical of networks is the most robust in the case of three component systems can be understood in general as resulting from the previously mentioned symmetry property of network robustness. Inspection of \ref{eq:sccrobustness} demonstrates that robustness is maximized by maximizing the number of edges \emph{between} the strongly connected components of the graph underlying system interdependencies.
% \section{Hierarchical systems are the most robust}
% \ref{fig:scc} shows systems with varying levels of hierarchy and modularity.

@@

\section{Stability criteria}
\todo{Since there is no point in publishing proofs of results contemporary
with Darwin's Origin of Species, this section will be replaced with a
reference to Gantmacher.}
Suppose that $M$ is an $n \times n$ matrix with real coefficients.
As described in \ref{sec:stabanalbn}, we call
the matrix $M$ stable if all its eigenvalues have negative real part.
Define the characteristic polynomial as $\chi(M)(z) = \det(zI - M)$.
Denote the coefficients of $\chi(M)$ as $s_j$ and its roots as $r_j$,
writing $p(x) = x^n + \sum_{k=1}^n s_k x^{n-k} = \prod_{k=1}^n (x-r_k)$.
Then, since the roots of $\chi$ are the eigenvalues of $M$, asking that
$M$ be stable is equivalent to asking that the roots of $\chi$ all have
negative real part.  Hence, let us call a polynomial stable if all its
roots have negative real part.  We will formulate conditions for this to
happen in terms of the coefficients of $\chi$.

@@

\section{Paths and eigenvalues}
\todo{I am thinking that we might want to jettison this section from
the published version and leave it as private notes for future work.
As a substitute for the published version, I made a section in front
of it which takes a more direct route to proving what is needed for
the purposes of showing that the characteristic polynomial factors and
making our arguments about total orders being the most stable.  We
should discuss this point tomorrow.}

@@

\todo{Start talking about Abelianizing the category and the algebra, say
more after decomposing into components.}

\todo{Characterize strongly connected components in terms of
fibration over poset.}

>>>>

% \section{Todo}
\listoftodos
\tableofcontents

@@

\section{Dynamical systems on biological networks}
In the construction of a class of potential models for biological systems at any level of the biological hierarchy from metabolic to ecosystem-level networks, it is common to first attempt to define a collection of observable phenomena of interest and determine a domain (such as binary numbers, integers, or real numbers) in which each observable can be quantified. Next, it is necessary to establish the interdependencies among system components. The specific manner in which the components depend upon one another must be clarified, which is often done by determining a particular system of mathematical functions that represents a hypothesis about how the quantified observables evolve in time. To the degree to which there is uncertainty about the interactions among system components, parameters are introduced to broaden the class of models under consideration. Finally, whatever model class remains may be compared to empirical observations to determine how capable the model is of representing the phenomena of interest.

For the case of continuous deterministic observables, the above process can be made more precise by associating a manifold $M_i$ to each observable, a directed graph, $G$, to the interdependencies, and a vector field $V \in \Gamma T(\prod_i M_i)$ over the space determined by taking the product of the manifolds associated to the collection of observables, satisfying these interdependencies for each observable  \cite{Deville}. For example, if we have two observables $\{x_1,x_2\}$ where the domains in which they are quantified are given by manifolds $\{M_1,M_2\}$ such that $x_1 \in M_1 \equiv \mathbb{R}^1$ and $x_2 \in M_2 \equiv \mathbb{R}^1$, a directed graph $G_X$ describing the dependencies between these observables and vector field $V$ with components $\{F_1,F_2\}$ defined on $M_1 \otimes M_2 \equiv \mathbb{R}^2$ satisfying the dependencies determined by $G_X$. If the system under consideration has the graph given in \ref{fig:examplesystemmodules}A
% \begin{center}
% \input{tex/examplehamdyngraph}
% \end{center}
with adjacency matrix
$$
\adj(G_X) = \begin{pmatrix}
0 & 1 \\
1 & 1
\end{pmatrix}
$$
having connectivity $\connectivity$ equal to the number of edges of the graph, in this case $\connectivity = 3$. For a general system the directed graph $G_X$ that describes the manner in which each of the variables depends upon one another is given by the adjacency matrix $\adj(G_X)$ where
 \begin{displaymath}
   \adj(G_X)_{ij} = \left\{
     \begin{array}{ll}
       1, & F_i \hbox{ depends on } x_j\\
       0, & F_i \hbox{ does not depend on } x_j
     \end{array}.
   \right.
\end{displaymath} For the system $X = \{G_X, M_X, F_X\}$, where $M_X \equiv \{M_1,M_2\}$ and $F_X \equiv \{F_1,F_2\}$ such that $F_1$ is constrained to be a function of $x_2$ and $F_2$ is a function of both $x_1$ and $x_2$ yielding the flow equations
\begin{align*}
\frac{dx_1}{dt} & = F_1(x_2)\\
\frac{dx_2}{dt} & = F_2(x_1,x_2)
\end{align*}

@@

\section{Stability analysis of biological networks}
Linearization of the $F_i$ about any fixed point $\vec{x}^0$ such that $F_i(\vec{x}^0)=0$ leads to the system
\begin{equation}\label{eq:lineardynsys}
\frac{d\vec{y}(t)}{dt} = A \vec{y}(t),
\end{equation}
where $\vec{y} = \vec{x} - \vec{x}^0$ and the $n \times n$-matrix $A$ has components
$$
a_{ij} = \left. \frac{\partial F_i}{\partial x_j} \right|_{\vec{x} = \vec{x}^0}.
$$
To each dynamical system having Jacobian matrix $A$ at some fixed point $\vec{x}^0$ we can associate a directed graph $G_A$ given by an adjacency matrix $\adj(G_A)$ where
 \begin{displaymath}
   \adj(G_A)_{ij} = \left\{
     \begin{array}{lr}
       1, & a_{ij} \neq 0\\
       0, & a_{ij} = 0
     \end{array}
   \right.
\end{displaymath}
The graphs $G_X$ and $G_A$ are equivalent because the condition $F_i$ independent of $x_j$ definitive of $\adj(G_X)$ corresponds precisely to $\frac{\partial F_i}{\partial x_j}=0$.

@@

\todo{Finish writing these criteria with reference to Gantmacher}
Define the characteristic polynomial as $\chi(M)(z) = \det(zI - M)$.
Denote the coefficients of $\chi(M)$ as $s_j$ and its roots as $r_j$,
writing $p(x) = x^n + \sum_{k=1}^n s_k x^{n-k} = \prod_{k=1}^n (x-r_k)$.
Then, since the roots of $\chi$ are the eigenvalues of $M$, asking that
$M$ be stable is equivalent to asking that the roots of $\chi$ all have
negative real part.  Hence, let us call a polynomial stable if all its
roots have negative real part.  We will formulate conditions for this to
happen in terms of the coefficients of $\chi$ \cite{Gantmacher1959}.

\section{Network hierarchy and strongly connected components}

A graph can be decomposed into a natural collection of \emph{modules} called strongly connected components. A strongly connected component of a graph is a maximal subset of vertices where each vertex within the subset can be reached from any other \cite{Cormen2009}. The strongly connected components of some examples of three component systems are outlined in \ref{fig:scc} along with their adjacency matrices and hierarchy diagrams (to be defined below).
\todo{add hierarchy diagrams to \ref{fig:scc}}

Corresponding to this decomposition we can construct a directed acyclic graph $\hier (G)$.  Each node of $\hier (G)$ corresponds to a strongly connected component of $G$. There is an edge from the node corresponding to component $C$ to the node corresponding to component $C'$ if and only if there exists a link from some vertex in $C$ to some vertex in $C'$ in $G$.  Because of the maximality of strongly connected components, $\hier (G)$ is acyclic.

One can also perform this construction in the oppsite direction.  Start with a directed acyclic graph $A$.  To each node $n$ of $A$ associate
a strongly connected graph $C_n$.  To each link $(i,j)$ of $A$ associate a non-empty subset of $\Vertex(C_i) \times \Vertex(C_j)$.  The result will be a graph $G$ such that $\hier(G) = A$ and furthermore, every graph $G$ such that $\hier(G) = A$ can be obtained in this manner.
\todo{explain how robustness in \ref{eq:robustness} is symmetric under permutations of the SCCs in $G$ or equivalently of the nodes of $\hier{G}$.}

\todo{add panel to \ref{fig:scc} showing that many graphs get mapped to the same SCC DAG}

\section{Stability and robustness analysis of particular system ensembles}
Suppose that $M$ is an $n \times n$ matrix with real coefficients and connectivity $\connectivity$. We will consider the case in which each of the parameters defining the matrix is sampled from the uniform distribution $\mathcal{U}(-1,1)$ so that the parameter space corresponds to the $\connectivity$-dimensional hypercube, $H^\connectivity$, of edge length $r=2$, centered about the origin, and having volume $r^\connectivity$.  We want to know the probability that, if we resample the elements $k_1, k_2, \ldots k_m$ of this matrix, the resulting matrix will correspond to a stable matrix.  This is a special case of \ref{eq:robustness}:
\begin{equation}\label{eq:condprobgen}
 R (\mathrm{stab}, k_1, \ldots k_m) =
  \frac{\int_{H^\connectivity} dx_1 \cdots dx_d \int_{H^m} {dx'}_{k_1} \cdots {dx'}_{k_m}\,
    \mathbf{1}_{stab} \mathbf{1}_{stab'}}
  {\int_{H^\connectivity} dx_1 \cdots dx_d  \, \mathbf{1}_{stab}(x_1 \cdots x_d)}
\end{equation}
where we made the abbreviations
\begin{align*}
\mathbf{1}_{stab} &= \mathbf{1}_{stab}(x_1 \cdots x_{k_1} \cdots x_{k_m} \cdots x_\connectivity) \\
\mathbf{1}_{stab'} &= \mathbf{1}_{stab'}(x_1 \cdots {x'}_{k_1} \cdots {x'}_{k_m}  \cdots x_d)
\end{align*}
i.e. $\mathbf{1}_{stab'}$ corresponds to replacing $x_{k_1}, \ldots x_{k_m}$ with their primed counterparts.

@@

@@

It has been stated previously on the basis of simulation that system stability decreases with connectivity as the system size goes to infinity \cite{May1972}. For small system sizes such as the two and three component systems, studied here the situation is not so clear cut. For two component systems, system stability is constant across the entire range of connectivities. For three component systems, the trend shows a minor decrease from connectivity $4$ to $5$ followed by small fluctuations as shown in \ref{fig:combined}A.
The relationship between connectivity and robustness for two component systems is shown in \ref{tab:structstabmat} and likewise for three component systems in \ref{tab:structstabmat3} and \ref{fig:combined}B. If we average over the different classes of matrices for a given connectivity we see there is a correlation between connectivity and robustness demonstrated by the red dots in \ref{fig:combined}B.
% \section{Cycle number is inversely correlated with robustness}
\ref{fig:combined}C shows the robustness for all three component systems as a function of the number of simple cycles (elementary circuits) of length greater than one in the corresponding directed graph \cite{Johnson1975}. There appears to be a weak negative correlation between robustness and the number of simple cycles.

@@

\section{Component decomposition, hierarchy, and the symmetries of robustness}

@@

Next, we ask what is the maximally robust graph that has a given set of connected components.
\begin{align}
R_{\mathrm{max}} = \frac{\frac{1}{2} ((\sum_i n_i)^2 - \sum_i n_i^2) + \sum_i c_i R_i}{\frac{1}{2} ((\sum_i n_i)^2 - \sum_i n_i^2) + \sum_i c_i}, \label{eq:sccmaxrobustness}
\end{align}
\todo{explain how robustness in \ref{eq:sccrobustness} is symmetric under transformations on the the SCCs}

% \section{Strongly connected components, robustness, and hierarchy}
The fact that the most hierarchical of networks is the most robust in the case of three component systems can be understood in general as resulting from the previously mentioned symmetry property of network robustness. Inspection of \ref{eq:sccrobustness} demonstrates that robustness is maximized by maximizing the number of edges \emph{between} the strongly connected components of the graph underlying system interdependencies.
% \section{Hierarchical systems are the most robust}
% \ref{fig:scc} shows systems with varying levels of hierarchy and modularity.

@@

\section{Stability criteria}
% \todo{Since there is no point in publishing proofs of results contemporary with Darwin's Origin of Species, this section will be replaced with a reference to Gantmacher.}
Suppose that $M$ is an $n \times n$ matrix with real coefficients.
As previously described, we call
the matrix $M$ stable if all its eigenvalues have negative real part.
Define the characteristic polynomial as $\chi(M)(z) = \det(zI - M)$.
Denote the coefficients of $\chi(M)$ as $s_j$ and its roots as $r_j$,
writing $p(x) = x^n + \sum_{k=1}^n s_k x^{n-k} = \prod_{k=1}^n (x-r_k)$.
Then, since the roots of $\chi$ are the eigenvalues of $M$, asking that
$M$ be stable is equivalent to asking that the roots of $\chi$ all have
negative real part.  Hence, let us call a polynomial stable if all its
roots have negative real part.  We will formulate conditions for this to
happen in terms of the coefficients of $\chi$.

@@

\section{Paths and eigenvalues}
% \todo{I am thinking that we might want to jettison this section from
% the published version and leave it as private notes for future work.
% As a substitute for the published version, I made a section in front
% of it which takes a more direct route to proving what is needed for
% the purposes of showing that the characteristic polynomial factors and
% making our arguments about total orders being the most stable.  We
% should discuss this point tomorrow.}

@@

% \todo{Start talking about Abelianizing the category and the algebra, say
% more after decomposing into components.}

% \todo{Characterize strongly connected components in terms of
% fibration over poset.}
